Mixture Models for Single-Cell Assays with Applications to Vaccine Studies

### 2013-06-25 MIMOSA/mimosafigures/manuscriptFigures.Rmd
Add code for generating manuscript figures

Read and merge the hvtn data. Cache it.

@@

read the fluidigm data

@@

Process it

@@

Fit the MIMOSA model to the HVTN data all cytokines and ENV-1-PTEG using EM and MCMC

```{r }
if(!file.exists("hvtnEMfit.rds")){
  hvtn.result.EM<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="EM",run.parallel=TRUE)
  saveRDS(hvtn.result.EM,file="hvtnEMfit.rds")
}else{
  hvtn.result.EM<-readRDS("hvtnEMfit.rds")
}
if(!file.exists("hvtnMCMCfit.rds")){
  hvtn.result.mcmc<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="mcmc",run.parallel=TRUE)
  saveRDS(hvtn.result.mcmc,file="hvtnMCMCfit.rds")
}else{
  hvtn.result.mcmc<-readRDS("hvtnMCMCfit.rds")
}
```

drop the stuff fit via mcmc and summarize, fit LRT, Fisher's and rank by log fold change

```{r cache=TRUE}
hvtn.summary.EM<-summarizeMIMOSAFit(hvtn.result.EM)
hvtn.summary.EM<-subset(hvtn.summary.EM,Method%in%"EM")
hvtn.summary.mcmc<-summarizeMIMOSAFit(hvtn.result.mcmc)
hvtn.summary.mcmc<-FisherFromSummary(hvtn.summary.mcmc)
hvtn.summary.mcmc<-LRT(hvtn.summary.mcmc)
hvtn.summary.mcmc<-FoldChangeRank(hvtn.summary.mcmc)
```

Compute ROC and FDR curves and calculate the AUC.
+ cache=TRUE

```{r }
ROC<-ComputeROCs(hvtn.summary.mcmc,hvtn.summary.EM)
FDR<-ComputeFDRs(hvtn.summary.mcmc,hvtn.summary.EM)
AUC<-ComputeAUCs(ROC)
```

FDR and ROC plots
--------
We see that MIMOSA performs as well or better than competing methods. The MCMC routine performs better than the EM routine for some data.

```{r cache=TRUE,fig.width=10,fig.height=8,fig.cap='The TNF expressing cytokine specific cells have high background as evidenced by the poor performance of all methods fit to TNF+ subsets of cells, with the exception of the triple-positive cell subset. As the number of simultaneously expressed cytokines being examined increases, the background decreases.'}
ROC$Method<-relevel(ROC$Method,"MIMOSA (EM)")
ggplot(subset(ROC,TCELLSUB%in%"cd3+/cd4+"))+geom_line(aes(x=FPR,y=TPR,color=Method))+theme_bw()+facet_wrap(~CYTOKINE)+geom_text(aes(x=x,y=y,label=sprintf("AUC=%s",signif(AUC,2)),col=Method),size=3,data=AUC,show_guide=FALSE)+theme(axis.text.x=element_text(angle=90,hjust=1))
```{r fig.width=10,fig.height=8,fig.cap='The FDR is well controlled for cytokine subsets with low background. The TNF expressing subsets have high background and consequently positive assay results in the "Visit 2" samples, which are labelled as "true negatives" in this scheme. The result is observed as a large observed false discovery rate for all methods.'}
FDR$Method<-relevel(FDR$Method,"MIMOSA (EM)")
ggplot(subset(FDR,TCELLSUB%in%c("cd3+/cd4+")))+geom_line(aes(y=true.fdr,x=fdr,color=Method))+geom_abline(1,lty=3)+theme_bw()+scale_x_continuous(name="Nominal FDR")+scale_y_continuous(name="Observed FDR")+facet_wrap(~CYTOKINE)+theme(axis.text.x=element_text(angle=90,hjust=1))
```

Process the fluidigm data to construct an expression set for MIMOSA
---------

@@

Fit MIMOSA by gene, grouping by stimulation and for all samples together.
----

@@

Heatmap of posterior probabilities of response for fit of all samples by gene.
----
-Red and orange are CMV (pp65). Yellow and green are HIV (gag and nef).
-When fitting the data to all samples simultaneously, we still observe clustering of samples by stimulation (i.e. CMV stimulated samples cluster together, and HIV stimulated samples cluster together).

```{r heatmap.probs.bygene,cache=TRUE}
foo<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)x@z[,2]))
props<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)do.call(c,lapply(x@result@p,function(x)-diff(x[,2])))))
S<-sign(props)
S[S==0]<-1
colnames(S)<-pData(fluidigm.fits.by.stim.allgenes[[1]][[1]])[,c("Patient.ID")]
stim<-pData(fluidigm.fits.by.stim.allgenes[[1]][[1]])$Stim
rownames(S)<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)as.character(unique(pData(x)$Gene))))
heatmap.2((S*foo)[apply(foo,1,function(x)sum(x>.1)>=2),],col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="signed Posterior Probabilities\n(fit by gene, all stimulations)")
```

Heatmap of posterior differences of proportions fit by gene for all samples.
----
-Red and orange are CMV (pp65). Yellow and green are HIV (gag and nef).
-When fitting the data to all samples simultaneously, we still observe clustering of samples by stimulation (i.e. CMV stimulated samples cluster together, and HIV stimulated samples cluster together).

```{r heatmap.props.bygene,cache=TRUE}
colnames(props)<-colnames(S)
rownames(props)<-rownames(S)
heatmap.2(S*sqrt(abs(props)),col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="Posterior Differences in Proportions\n(fit by gene all stimulations)")
```

Heatmap of posterior probability of response for fit of samples by stimulation and by gene
---
-Red and orange are CMV (pp65). Yellow and green are HIV (gag and nef).

```{r heatmap.probs.bystim.bygene,cache=TRUE}
foo<-do.call(cbind,lapply(fluidigm.fits.by.stim.bygene,function(x)do.call(rbind,lapply(x,function(x)x@z[,2]))))
props<-do.call(cbind,lapply(fluidigm.fits.by.stim.bygene,function(x)do.call(rbind,lapply(x,function(x)do.call(c,lapply(x@result@p,function(x)-diff(x[,2])))))))
S<-sign(props)
colnames(S)<-do.call(c,lapply(fluidigm.fits.by.stim.bygene,function(x)as.character(pData(x[[1]])[,c("Patient.ID")])))
S[S==0]<-1
stim<-as.numeric(factor(do.call(c,lapply(fluidigm.fits.by.stim.bygene,function(x)as.character(pData(x[[1]])$Stim)))))
colnames(S)<-do.call(c,lapply(fluidigm.fits.by.stim.bygene,function(x)as.character(pData(x[[1]])$Patient.ID)))
rownames(S)<-do.call(cbind,lapply(fluidigm.fits.by.stim.bygene[[1]],function(x)as.character(unique(pData(x)$Gene))))
heatmap.2((S*foo)[apply(foo,1,function(x)sum(x>.1)>=1),],col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],main="Signed Posterior Probabilities\n(fit by gene and stimulation).")
```

Heatmap of posterior differences in proportions for fit of samples by stimulation and by gene. 
---
-Red and orange are CMV (pp65). Yellow and green are HIV (gag and nef).

```{r heatmap.props.bystim.bygene,cache=TRUE}
colnames(props)<-colnames(S)
rownames(props)<-rownames(S)
heatmap.2(sqrt(abs(props))*S,col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="Posterior differences in proportions\n(fit by gene and stimulations)")
```

Fisher's exact test and heatmaps of -log10(q-values) from Fisher's
---
There is little clustering by stimulation evident from testing for differential expression with Fisher's exact test.

```{r cache=TRUE}
fisher<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x){
  p<-vector('numeric',nrow(x@result@n.stim))
  for(i in 1:nrow(x@result@n.stim)){
    p[i]<-fisher.test(as.matrix(rbind(x@result@n.stim[i,],x@result@n.unstim[i,])),alternative="two.sided")$p.value
  }
  p
}))
S<-sign(lapply(fluidigm.fits.by.stim.allgenes,function(x)do.call(rbind,lapply(x,function(x)(prop.table(as.matrix(x@result@n.stim),1)-prop.table(as.matrix(x@result@n.unstim),1))[,2])))[[1]])
S[S==0]<-1
colnames(fisher)<-gsub("_B3GAT1_Treatment","",colnames(S))
rownames(fisher)<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)as.character(unique(pData(x)$Gene))))
stim<-pData(fluidigm.fits.by.stim.allgenes[[1]][[1]])$Stim
heatmap.2(fisher*S,col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="-log10(p-values)\nfrom Fisher's Exact Test")
```

Multivariate MIMOSA on fluidigm data looking at interactions
---------
There is a difference when comparing stimulations and looking at the multivariate vector of expression for GZMK and CCR7 (i.e [+/+,+/-,-/+,--]), but no difference is detected when we look at expression of GZMK | CCR7, since one cell subset increases while the other decreases upon stimultion.

```{r cache=FALSE}
fl.comb<-combine(fluidigm[[1]],fluidigm[[2]],fluidigm[[3]],fluidigm[[4]])
melted<-melt(fl.comb)
require(glmnet)
require(Matrix)
melted$Et.bin<-as.numeric(as.logical(melted$Et))
#setnames(melted,"__wellKey","wellKey")
melted<-within(melted,group<-Patient.ID:Stim.Agent)
melted<-cast(melt(melted,id=c("group","Stim.Condition","primerid","wellKey"),measure="Et.bin"),group+Stim.Condition+wellKey~primerid)

interact.scan<-combn(colnames(melted)[4:99],2,function(x){
  g1<-x[1];g2<-x[2]
  gg<-paste(g1,g2,sep=":")
  bar<-data.frame(melted[,c("group","Stim.Condition",g1,g2)])
  bar$foo<-factor(bar[,c(g1)]):factor(bar[,c(g2)])
  setnames(bar,"foo",gg)
  bar[,5]<-factor(bar[,5],levels=c("0:0","0:1","1:0","1:1"))
  C<-cast(melt(bar,id=c("group","Stim.Condition"),measure=c(gg)),group+Stim.Condition~value,fun.aggregate=function(x)length(x))
  expected<-c("0:0","0:1","1:0","1:1")
  unused<-setdiff(expected,colnames(C))
  newframe<-NULL
  for(i in seq_along(expected)){
    if(expected[i]%in%unused){
      newframe<-cbind(newframe,rep(0,nrow(C)))
    }else{
      newframe<-cbind(newframe,C[,expected[i]])
    }
  }
  colnames(newframe)<-expected
  C<-cbind(C[,1:2],newframe)
  CC<-list(n.unstim=subset(C,Stim.Condition%in%"Unstim")[,-c(1:2)],n.stim=subset(C,Stim.Condition%in%"Stim")[,-c(1:2)])
  foo<-lapply(CC,function(x)prop.table(as.matrix(x),1))
  psum<-mean(rowSums(foo[[1]][,2:4])-rowSums(foo[[2]][,2:4]))
  pij<-colMeans(foo[[1]][,2:4]-foo[[2]][,2:4])
  res<-c(psum,pij)
  res
})

genes<-combn(colnames(melted)[6:101],2)[,which(abs(interact.scan[1,])<0.01&abs(interact.scan[4,])>0.3)]
interact.scan[,which(abs(interact.scan[1,])<0.01&abs(interact.scan[4,])>0.3)]
ggg<-genes[,6]
g1<-ggg[1];g2<-ggg[2] #CCR7 and GZMK
g1<-"CCR7";g2<-"GZMK"
gg<-paste(g1,g2,sep=":")
bar<-data.frame(melted[,c("group","Stim.Condition",g1,g2)])
bar$foo<-factor(bar[,c(g1)]):factor(bar[,c(g2)])
setnames(bar,"foo",gg)
C<-cast(melt(bar,id=c("group","Stim.Condition"),measure=c(gg)),group+Stim.Condition~value,fun.aggregate=function(x)length(x))
CC<-list(n.unstim=subset(C,Stim.Condition%in%"Unstim")[,-c(1:2)],n.stim=subset(C,Stim.Condition%in%"Stim")[,-c(1:2)])

toplot<-data.frame(rbind(data.frame(CC$n.stim),data.frame(CC$n.unstim)),stim=c(rep("Stim",nrow(CC$n.stim)),rep("Unstim",nrow(CC$n.unstim))))
toplot$id<-rep(unlist(lapply(strsplit(as.character(subset(C,Stim.Condition%in%"Unstim")[,1]),":"),function(x)x[1]),use.names=FALSE),2)
colnames(toplot)[1:4]<-gsub("X","",paste(g1,g2,colnames(toplot)[1:4],sep=":"))
toplot<-melt(toplot,id=c("id","stim"))
ggplot(toplot)+geom_raster(aes(x=variable,y=id,fill=value))+theme_bw()+theme(axis.text.x=element_text(angle=90))+facet_wrap(~stim)+scale_fill_continuous(na.value="gray",breaks=round(seq(0,96,l=10)))

DD<-CC
DD[[1]]<-cbind(CC[[1]][,1],rowSums(CC[[1]][,2:4]))
DD[[2]]<-cbind(CC[[2]][,1],rowSums(CC[[2]][,2:4]))
toplot2<-data.frame(rbind(data.frame(DD$n.stim),data.frame(DD$n.unstim)),stim=c(rep("Stim",nrow(DD$n.stim)),rep("Unstim",nrow(DD$n.unstim))))
toplot2$id<-rep(unlist(lapply(strsplit(as.character(subset(C,Stim.Condition%in%"Unstim")[,1]),":"),function(x)x[1]),use.names=FALSE),2)
colnames(toplot2)[1:2]<-gsub("X","",paste(g1,g2,colnames(toplot2)[1:2],sep=":"))
toplot2<-melt(toplot2,id=c("id","stim"))
ggplot(toplot2)+geom_raster(aes(x=variable,y=id,fill=value))+theme_bw()+theme(axis.text.x=element_text(angle=90))+facet_wrap(~stim)+scale_fill_continuous(na.value="gray",breaks=round(seq(0,96,l=10)))

>>>>

### Read and merge the hvtn data.

@@

### Read the fluidigm data.

@@

Process it.

@@

### Fit the MIMOSA model to the HVTN data all cytokines and ENV-1-PTEG using EM and MCMC implementations.

```{r }
if(!file.exists("hvtnEMfit.rds")){
  hvtn.result.EM<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="EM",run.parallel=TRUE)
  saveRDS(hvtn.result.EM,file="hvtnEMfit.rds")
}else{
  hvtn.result.EM<-readRDS("hvtnEMfit.rds")
}
if(!file.exists("hvtnMCMCfit.rds")){
  hvtn.result.mcmc<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="mcmc",run.parallel=TRUE,pXi=0.1,getP=TRUE)
  saveRDS(hvtn.result.mcmc,file="hvtnMCMCfit.rds")
}else{
  hvtn.result.mcmc<-readRDS("hvtnMCMCfit.rds")
}
```

### Summarize the results. Do the likelihood ratio test, Fisher's test, and rank by log fold change for comparison.

```{r cache=TRUE,message=FALSE,warning=FALSE}
hvtn.summary.EM<-summarizeMIMOSAFit(hvtn.result.EM)
hvtn.summary.EM<-subset(hvtn.summary.EM,Method%in%"EM")
hvtn.summary.mcmc<-summarizeMIMOSAFit(hvtn.result.mcmc)
hvtn.summary.mcmc<-FisherFromSummary(hvtn.summary.mcmc)
hvtn.summary.mcmc<-LRT(hvtn.summary.mcmc)
hvtn.summary.mcmc<-FoldChangeRank(hvtn.summary.mcmc)
```

### Compute ROC and FDR curves and calculate the AUC for all the results.

```{r cache=TRUE}
ROC<-ComputeROCs(hvtn.summary.mcmc,hvtn.summary.EM)
FDR<-ComputeFDRs(hvtn.summary.mcmc,hvtn.summary.EM)
AUC<-ComputeAUCs(ROC)
```

### FDR and ROC plots of the ICS data.
--------
We see that MIMOSA performs as well or better than competing methods. The MCMC routine performs better than the EM routine for some data.

```{r cache=TRUE,fig.width=10,fig.height=8,fig.cap='The TNF expressing cytokine specific cells have high background as evidenced by the poor performance of all methods fit to TNF+ subsets of cells, with the exception of the triple-positive cell subset. As the number of simultaneously expressed cytokines being examined increases, the background decreases.'}
ROC$Method<-relevel(ROC$Method,"MIMOSA (EM)")
levs <-
  c("IFN-IL2-TNF+", "IFN-IL2+TNF-", "IFN-IL2+TNF+", "IFN+IL2-TNF-", 
    "IFN+IL2-TNF+", "IFN+IL2+TNF-", "IFN+IL2+TNF+", "IFNg+", "IFNg+IL2+", 
    "IFNg+TNF+", "IL2+", "IL2-IFNg+", "IL2+ OR IFNg+", "IL2+IFNg-", 
    "IL2+IFNg+", "IL2+TNF+", "TNF+")
ROC$CYTOKINE<-factor(ROC$CYTOKINE,labels=levs)
AUC$CYTOKINE<-factor(AUC$CYTOKINE,labels=levs)
rocplots<-ggplot(subset(ROC,TCELLSUB%in%"cd3+/cd4+"))+geom_line(aes(x=FPR,y=TPR,color=Method))+theme_bw()+facet_wrap(~CYTOKINE)+geom_text(aes(x=x,y=y,label=sprintf("AUC=%s",signif(AUC,2)),col=Method),size=3,data=AUC,show_guide=FALSE)+theme(axis.text.x=element_text(angle=90,hjust=1))
```{r fig.width=10,fig.height=8,fig.cap='The FDR is well controlled for cytokine subsets with low background. The TNF expressing subsets have high background and consequently positive assay results in the "Visit 2" samples, which are labelled as "true negatives" in this scheme. The result is observed as a large observed false discovery rate for all methods.'}
FDR$Method<-relevel(FDR$Method,"MIMOSA (EM)")
levs<-levels(FDR$CYTOKINE)
levs <-
  c("IFN-IL2-TNF+", "IFN-IL2+TNF-", "IFN-IL2+TNF+", "IFN+IL2-TNF-", 
    "IFN+IL2-TNF+", "IFN+IL2+TNF-", "IFN+IL2+TNF+", "IFNg+", "IFNg+IL2+", 
    "IFNg+TNF+", "IL2+", "IL2-IFNg+", "IL2+ OR IFNg+", "IL2+IFNg-", 
    "IL2+IFNg+", "IL2+TNF+", "TNF+")
FDR$CYTOKINE<-factor(FDR$CYTOKINE,labels=levs)
fdrplots<-ggplot(subset(FDR,TCELLSUB%in%c("cd3+/cd4+")))+geom_line(aes(y=true.fdr,x=fdr,color=Method))+geom_abline(1,lty=3)+theme_bw()+scale_x_continuous(name="Nominal FDR")+scale_y_continuous(name="Observed FDR")+facet_wrap(~CYTOKINE)+theme(axis.text.x=element_text(angle=90,hjust=1))
rocplots
fdrplots
```

### Figure 1 and S1.

```{r echo=FALSE,eval=FALSE}
suppressPackageStartupMessages(library(gridExtra))
f1levs<-levs[c(8,11,13,17,15,16,10,7)]
sflevs<-setdiff(levs,f1levs)
AUC$y[AUC$Method%in%"Fisher"]<-0.4
AUC$y[AUC$Method%in%"MIMOSA (EM)"]<-0.5
AUC$y[AUC$Method%in%"MIMOSA (MCMC)"]<-0.3

fnames<-c("revised_2.pdf","revised_1.pdf","revised_15.pdf","revised_3.pdf","revised_5.pdf","revised_6.pdf","revised_7.pdf","revised_8.pdf")
for(i in seq_along(f1levs)){
  fi.1<-ggplot(subset(ROC,TCELLSUB%in%"cd3+/cd4+"&CYTOKINE%in%f1levs[i]))+geom_line(aes(x=FPR,y=TPR,color=Method))+theme_bw()+facet_wrap(~CYTOKINE)+geom_text(aes(x=x,y=y,label=sprintf("AUC=%s",signif(AUC,2)),col=factor(Method)),data=subset(AUC,CYTOKINE%in%f1levs[i]),show_guide=FALSE)+theme(axis.text.x=element_text(angle=90,hjust=1)) 
  fi.2<-ggplot(subset(FDR,TCELLSUB%in%c("cd3+/cd4+")&CYTOKINE%in%f1levs[i]))+geom_line(aes(y=true.fdr,x=fdr,color=Method))+geom_abline(1,lty=3)+theme_bw()+scale_x_continuous(name="Nominal FDR")+scale_y_continuous(name="Observed FDR")+facet_wrap(~CYTOKINE)+theme(axis.text.x=element_text(angle=90,hjust=1))
  pdf(file=sprintf("/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/%s",fnames[i]),width=10,height=3)
  grid.arrange(fi.1,fi.2,ncol=2,nrow=1)
  dev.off()
  grid.arrange(fi.1,fi.2,ncol=2,nrow=1)
}
```

### Process the fluidigm data to construct an expression set for MIMOSA.
---------

@@

### Fit MIMOSA by gene, grouping by stimulation and for all samples together.
----

@@

### Heatmaps for Fluidigm data fit to all stimulations simultaneously.
----
-Red and orange are CMV (pp65). Yellow and green are HIV (gag and nef).
-When fitting the data to all samples simultaneously, we still observe clustering of samples by stimulation (i.e. CMV stimulated samples cluster together, and HIV stimulated samples cluster together).

```{r heatmap.probs.bygene,cache=TRUE}
foo<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)x@z[,2]))
props<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)do.call(c,lapply(x@result@p,function(x)-diff(x[,2])))))
S<-sign(props)
S[S==0]<-1
colnames(S)<-pData(fluidigm.fits.by.stim.allgenes[[1]][[1]])[,c("Patient.ID")]
stim<-pData(fluidigm.fits.by.stim.allgenes[[1]][[1]])$Stim
rownames(S)<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)as.character(unique(pData(x)$Gene))))
#heatmap.2((S*(foo))[apply(foo,1,function(x)sum(x>0.6)>=1),],col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="signed Posterior Probabilities\n(fit by gene, all stimulations)")
foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
```

### Figure 2 B, (fit of all samples simultaneously)

```{r eval=FALSE,echo=FALSE}
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/FluidigmCombinedFit.pdf")
foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
dev.off()
```

Heatmap of posterior differences in proportions from simultaneous fit (not shown).
----
-Red and orange are CMV (pp65). Yellow and green are HIV (gag and nef).
-When fitting the data to all samples simultaneously, we still observe clustering of samples by stimulation (i.e. CMV stimulated samples cluster together, and HIV stimulated samples cluster together).

```{r heatmap.props.bygene,cache=TRUE,include=FALSE}
colnames(props)<-colnames(S)
rownames(props)<-rownames(S)
heatmap.2(S*sqrt(abs(props)),col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="Posterior Differences in Proportions\n(fit by gene all stimulations)")
```

### Heatmaps for Fluidigm data fit per stimulation.
---
-Red and orange are CMV (pp65). Yellow and green are HIV (gag and nef).

```{r heatmap.probs.bystim.bygene,cache=TRUE}
foo<-do.call(cbind,lapply(fluidigm.fits.by.stim.bygene,function(x)do.call(rbind,lapply(x,function(x)x@z[,2]))))
props<-do.call(cbind,lapply(fluidigm.fits.by.stim.bygene,function(x)do.call(rbind,lapply(x,function(x)do.call(c,lapply(x@result@p,function(x)-diff(x[,2])))))))
S<-sign(props)
colnames(S)<-do.call(c,lapply(fluidigm.fits.by.stim.bygene,function(x)as.character(pData(x[[1]])[,c("Patient.ID")])))
S[S==0]<-1
stim<-as.numeric(factor(do.call(c,lapply(fluidigm.fits.by.stim.bygene,function(x)as.character(pData(x[[1]])$Stim)))))
colnames(S)<-do.call(c,lapply(fluidigm.fits.by.stim.bygene,function(x)as.character(pData(x[[1]])$Patient.ID)))
rownames(S)<-do.call(cbind,lapply(fluidigm.fits.by.stim.bygene[[1]],function(x)as.character(unique(pData(x)$Gene))))
#h<-heatmap.2((S*foo)[apply(foo,1,function(x)sum(x>.6)>=1),],col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],main="Signed Posterior Probabilities\n(fit by gene and stimulation).")
foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
h<-heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
hmaporder<-h$colInd
geneinds<-apply(foo,1,function(x)sum(x>.6)>=1)
```

### Figure 2 A (fit per stimulation)
Fit MIMOSA per stimulation.

```{r eval=FALSE,echo=FALSE}
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Fluidigm_plot1.pdf")
foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
dev.off()
```

Heatmap of posterior differences in proportions for fit of samples by stimulation and by gene (not shown).
---
-Red and orange are CMV (pp65). Yellow and green are HIV (gag and nef).

```{r heatmap.props.bystim.bygene,cache=TRUE,include=FALSE}
colnames(props)<-colnames(S)
rownames(props)<-rownames(S)
heatmap.2(sqrt(abs(props))*S,col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="Posterior differences in proportions\n(fit by gene and stimulations)")
```

### Fisher's exact test and heatmaps of -log10(q-values) from Fisher's test.
---
There is little clustering by stimulation evident from testing for differential expression with Fisher's exact test.

```{r cache=TRUE}
fisher<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x){
  p<-vector('numeric',nrow(x@result@n.stim))
  for(i in 1:nrow(x@result@n.stim)){
    p[i]<-fisher.test(as.matrix(rbind(x@result@n.stim[i,],x@result@n.unstim[i,])),alternative="two.sided")$p.value
  }
  p
}))
adj.fisher<-matrix(p.adjust(fisher,"fdr"),nrow=96)
S<-sign(lapply(fluidigm.fits.by.stim.allgenes,function(x)do.call(rbind,lapply(x,function(x)(prop.table(as.matrix(x@result@n.stim),1)-prop.table(as.matrix(x@result@n.unstim),1))[,2])))[[1]])
S[S==0]<-1
colnames(fisher)<-gsub("_B3GAT1_Treatment","",colnames(S))
rownames(fisher)<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)as.character(unique(pData(x)$Gene))))
stim<-pData(fluidigm.fits.by.stim.allgenes[[1]][[1]])$Stim
#heatmap.2(fisher*S,col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="-log10(p-values)\nfrom Fisher's Exact Test")
colnames(adj.fisher)<-colnames(fisher)
rownames(adj.fisher)<-rownames(fisher)
heatmap.2((((adj.fisher))*S)[apply(adj.fisher,1,function(x)sum(x<0.1)>1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
```

### Figure 2 D (q-values from Fisher's test)

```{r eval=FALSE,echo=FALSE}
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/FluidigmFisher.pdf")
colnames(adj.fisher)<-colnames(fisher)
rownames(adj.fisher)<-rownames(fisher)
heatmap.2((((adj.fisher))*S)[apply(adj.fisher,1,function(x)sum(x<0.1)>1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
dev.off()

```{r eval=TRUE,echo=FALSE}
emprop<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x){
  p<-vector('numeric',nrow(x@result@n.stim))
  p<-(prop.table(as.matrix((x@result@n.stim)),1)-prop.table(as.matrix(x@result@n.unstim),1))[,2]
}))
rownames(emprop)<-rownames(fisher)
colnames(emprop)<-colnames(fisher)

```{r echo=FALSE,warning=FALSE}
heatmap.2(emprop[geneinds,hmaporder],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),Colv=NULL,denscol="black",tracecol="black",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)[hmaporder]])
```

### Figure 2C empirical proportions.

```{r eval=FALSE}
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/FluidigmEmpiricalProps.pdf")
heatmap.2(emprop[geneinds,hmaporder],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),Colv=NULL,denscol="black",tracecol="black",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)[hmaporder]])
dev.off()
```
